A weakly structured stem for human origins in Africa

### 2023-02-27 paper.tex
add some methods

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (SI Section
\ref{sec:models}). These parameters were ones supported by multiple genetic and
archaeological studies. Fixed parameters included the time of divergence
between western and eastern African populations, set to $60$ka, just prior to
the split of Eurasians and East Africans set to $50$ka. We also fixed the
amount of admixture from Neanderthals to Europeans directly following the
out-of-Africa migration to $1.5\%$ at $45$ka.

@@

\aprcomment{This can be very short}

>>>>

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (SI Section
\ref{sec:models}). These parameters were ones supported by multiple genetic and
archaeological studies \citep{Bergstrom2021-iw}.
Fixed parameters included the time of divergence
between western and eastern African populations, set to $60$ka, just prior to
the split of Eurasians and East Africans set to $50$ka. We also fixed the
amount of admixture from Neanderthals to Europeans directly following the
out-of-Africa migration to $1.5\%$ at $45$ka.

@@

\subsection*{Data and sequencing}

We used a subset of populations from the merged sequencing data from the
African Diversity Reference Panel (ADRP) \citep{Gurdasani2015-qy,Pagani2015-pz}
and the Thousand Genomes Project (1KGP)
\citep{1000_Genomes_Project_Consortium2015-zq}. These included the Amhara,
Bakiga, Gumuz, Nama, Oromo, Somali, and Zulu from ADRP, and ESN, GWD, LWK, MSL,
YRI, CEU, GBR, CHB, and PJL from 1KGP. After filtering for relatedness and
retaining Nama individuals with $>70\%$ estimated Khoe-San ancestry, we used
data from 289 individuals, including 44 Nama individuals. These were merged
with the high-coverage Neanderthal genome from Vindija Cave
\citep{Prufer2017-kk}. We kept variants from regions that fell within the 1KGP
srict callability mask, overlapped with at least 100 continuously called base
pairs in the Neanderthal genome, and were annotated as intergenic.
\texttt{ADMIXTURE} and PCA analyses were conducted on a subset of variants
filtered to remove variants in high LD ($r^2$ threshold of $0.1$). See SI
section \ref{sec:data-sequencing} for additional details on data and
sequencing.

\subsection*{LD and diversity statistics}

We used multi-population LD and pairwise diversity statistics to fit
parameterized demographic models to data \citep{Ragsdale2019-nt}. Unbiased LD
statistics were computed from all variants in retained intergenic regions
\citep{Ragsdale2020-nz}, for pairs of variants separated by recombination
distances $r = 5\times10^{-6}$ to $5\times10^{-3}$. These were assigned to 16
recombination distance bins, and average statistics were computed within each
bin (SI section \ref{sec:computing-stats}). Expected statistics were computed
in \texttt{moments}, which also performed likelihood-based parameter
optimization. Conditional site frequency spectra (cSFS) were computed
conditioned on the Vindija Neanderthal carrying the derived allele relative to
the ancestral allele determined by a six-primate alignment
\citep{1000_Genomes_Project_Consortium2015-zq}.

\subsection*{Model specification and fitting}

Model parameters include population sizes and size changes, split times,
continuous migration rates, and admixture times and proportions. The simplest
model we tested was a bifurcating tree-like structure, allowing for subsequent
migrations and recent known admixture events. To include ancestral population
structure, we tested models that included multiple stem groups, which each were
allowed their own population size and could be connected by continuous
migration. We tested multiple scenarios of early population structure,
including long-lasting continuous migration between stem populations and
scenarios of periods of isolation with subsequent merger events (SI
section~\ref{sec:modelspec}). To avoid overfitting, we to incrementally added
copmlexity to our model optimization, and we fixed a number of parameters that
are constrained by historical records or are consistently estimated across
multiple models and previous studies (SI section~\ref{sec:models}). Likelihoods
were computed using a composite multivariate Gaussian likelihood approach, and
confidence intervals were estimated by refitting each model to 200
block-bootstrap replicate datasets (SI
section~\ref{sec:likelihoods-uncertainty}). We iteratively used gradient
descent and L-BFGS-B optimization routines to fit each parameterized model (SI
section~\ref{sec:optimization}).

\subsection*{Gene genalogy reconstruction}

\subsection*{Conditional site frequency spectrum}
